Inferring cosmology from gravitational waves using non-parametric detector-frame mass distribution

### 2025-01-09 src/tex/ms.tex
[showyourwork] overleaf sync

\usepackage{showyourwork}
\usepackage{amsfonts,amssymb,amsmath}
\usepackage[nolist,nohyperlinks]{acronym}
\usepackage{bookmark}
\usepackage{xcolor}
\usepackage{array}
\usepackage{graphicx}
\usepackage{mathtools}
\usepackage{orcidlink}

@@

\begin{table}[htbp]
    \caption{Notation used in this paper}
    \begin{tabular}{cp{0.7\linewidth}}
        \toprule
        Notation & Description \\
        \midrule
        $m_1$ & Primary mass of the \ac{GW} sources in the source frame \\
        $q$ & Mass ratio of the \ac{GW} sources \\
        $z$ & Redshift of the \ac{GW} sources \\
        $d_L$ & Luminosity distance of the \ac{GW} sources \\
        $m^z_1$ & Primary mass of the \ac{BBH} sources in the detector frame \\
        $Y_t$ & detector-frame primary mass posterior probability distribution samples of the $t$-th \ac{GW} event \\
        $\mathbf{Y}$ & Set of detector-frame primary mass posterior probability distribution samples of all \ac{GW} events \\
        $\Theta_i$ & Hyperparameters of the \ac{(H)DPGMM} \\
        $\mathbf{\Theta}$ & Set of hyperparameters of the \ac{(H)DPGMM} \\
        $p(m^z_1|\mathbf{\Theta})$ & Detector-frame observed primary mass population distribution reconstructed with the \ac{(H)DPGMM} \\
        $\Lambda$ & Source-frame population model \\
        $\Omega$ & Cosmological model \\
        $p(m_1|\Lambda)$ & Source-frame primary mass population distribution assuming the population model $\Lambda$ \\
        $p(m^z_1|\Lambda, \Omega)$ & Detector-frame primary mass population distribution assuming the population model $\Lambda$ and cosmological model $\Omega$ \\
        $\mathrm{det}$ & detectability of the \ac{GW} events \\
        \botrule
    \end{tabular}
    \label{tab:notation}
\end{table}

@@

In this example, instead of transforming the source-frame population model to the detector frame for each $H_0$ and performing a grid search, we define a function that calculates $d_\mathrm{JS}$ for a given set of the free parameters.
We then use the optimization algorithm mentioned in Sec.~\ref{sec:remapping} to find the optimized free parameters that minimize $d_\mathrm{JS}$.
The optimization method requires bounds for the free parameters, which we show in Table~\ref{tab:bounds}.
\begin{table}[htbp]
    \caption{Free parameter bounds for the joint inference presented in Section~\ref{sec:inference_multi}.}
    \begin{tabular}{cc}
        \toprule
        Parameter & Bounds \\
        \midrule
        $H_0$ & $[1, 350]$ \\
        $\alpha$ & $[1.01, 15]$ \\
        $\mu$ & $[0.01, 70]$ \\
        $\sigma$ & $[0.01, 60]$ \\
        \botrule
    \end{tabular}
    \label{tab:bounds}
\end{table}
The initial guesses for the optimization are chosen randomly from a uniform distribution within the bounds.
The result of the remapping is shown in Fig.~\ref{fig:simulation_result_multi}.
\begin{figure}
    \script{plot_simulation_result_multi.py}
    \includegraphics[width=\linewidth]{figures/simulation_result_multi.pdf}
    \caption{
        Optimized parameter distribution for the mock data inference.
        Simulated values are marked in red.
    }
    \label{fig:simulation_result_multi}
\end{figure}
The recovered distribution, whose uncertainty budget comes from the remapping of the individual-\ac{PE} uncertainty in the non-parametric reconstruction, is consistent with the simulated values.

@@

Next, we chose the \textsc{power law + peak} model as the source-frame population model $\Lambda$ and the $\Lambda$CDM model as the cosmological model $\Omega$.
For the redshift population model, instead of the uniform distribution over comoving volume and source-frame time, we used the \textsc{power law} redshift evolution model described in Eq.~8 of \cite{KAGRA:2021duu}.
Similar to Sec.~\ref{sec:inference_multi}, we fixed some parameters of the population model and the cosmological model.
For the cosmological model, we fixed the parameters to the results from \cite{Planck:2018vyg} except $H_0$.
For the population model, we fixed $\alpha$, $w$, $m_\mathrm{min}$, $m_\mathrm{max}$, and $\delta$ to the results from \cite{LIGOScientific:2021aug}.
Note that the fixed parameters are not necessarily the true values, but they can be considered as a reasonable guess for the parameters.
The fixed parameters of the population model are shown in Table~\ref{tab:fixed_parameters}.
\begin{table}[htbp]
    \caption{Fixed parameters of the $\Lambda$CDM model (left) and the \textsc{power law + peak} model (right).}
    \begin{tabular}{cc|cc}
        \toprule
        Parameter & Value & Parameter & Value  \\
        \midrule
         & & $w$ & $0.024$ \\
        $\Omega_m$ & $0.315$ & $\alpha$ & $4.2\ \mathrm{M}_\odot$\\
        $\Omega_\Lambda$ & $0.685$ & $m_\mathrm{min}$ & $5\ \mathrm{M}_\odot$ \\
        $w$ & $-1$ & $m_\mathrm{max}$ & $109\ \mathrm{M}_\odot$ \\
         & & $\delta$ & $4.9\ \mathrm{M}_\odot$\\
        \botrule
    \end{tabular}
    \label{tab:fixed_parameters}
\end{table}

We then perform the remaining steps in the same way as Sec.~\ref{sec:inference_multi}.
We set the bounds for the power-law index of the redshift population model $\kappa$ to $[-100, 100]$, and the bounds for other free parameters are the same as in Table~\ref{tab:bounds}.
The result of the remapping is shown in Fig.~\ref{fig:real_result_multi}, and it is consistent with the posterior distribution reported in \cite{LIGOScientific:2021aug}.
\begin{figure}
    \script{plot_real_result_multi.py}
    \includegraphics[width=\linewidth]{figures/real_result_multi.pdf}
    \caption{
        Optimized parameter distribution for the real data inference.
    }
    \label{fig:real_result_multi}
\end{figure}
we find \variable{output/real_H0.txt}, (68\% credible interval around the median). Nonetheless, Figure~\ref{fig:real_result_multi} shows a peak in the marginalized posterior distribution of $H_0$ around $140\, \mathrm{km/s/Mpc}$, with a width of around $30-35\, \mathrm{km/s/Mpc}$ roughly consistent with the uncertainty reported in \cite{LIGOScientific:2021aug} ($H_0 = 50^{+37}_{-30} \, \mathrm{km/s/Mpc}$).
%This means a significant fraction of the reconstructed observed population distribution has an optimized $H_0$ around $140\, \mathrm{km/s/Mpc}$, which does not include the uncertainty from the mismatch between the population model and the observed population distribution.

The reason for the peak differing from the one reported in \cite{LIGOScientific:2021aug}, which is around $50\, \mathrm{km/s/Mpc}$, can be several, and most likely stemming from the various approximations made in this work. First of all, the choice of the Jensen-Shannon distance as a measure of ``closeness'' between the non-parametric reconstruction and the predicted population is an arbitrary choice and, despite being a commonly used metric, does not stem from a specific derivation and/or assumption on the process that generated the observations. As we stated in Section~\ref{sec:remapping}, the output of our method is not a posterior distribution in the strict sense due to this choice: the derivation of a well-grounded metric to connect non-parametric and parametric distributions will be discussed in a future paper. Another possibility is the approximation made while including the selection effects: as discussed in \cite{Essick:2023upv}, this choice can lead, to some extent, to a bias in the reconstruction, and thus be the reason why the peak of the posterior distribution for $H_0$ is shifted. Accounting for selection effects while reconstructing the observed distribution is in principle possible -- although not explored in the currently available literature -- and will be, as well, investigated in the future.

>>>>

\usepackage{showyourwork}
\usepackage{amsfonts,amssymb,amsmath}
\usepackage[nolist,nohyperlinks]{acronym}
\usepackage{bookmark}
\usepackage{xcolor}
\usepackage{array}
\usepackage{graphicx}
\usepackage{mathtools}
\usepackage{orcidlink}
\usepackage{tabularx}

@@

\begin{table}
    \caption{Notation used in this paper}
    \begin{tabular}{cp{0.7\linewidth}}
        \toprule
        Notation & Description \\
        \midrule
        $m_1$ & Primary mass of the \ac{GW} sources in the source frame \\
        $q$ & Mass ratio of the \ac{GW} sources \\
        $z$ & Redshift of the \ac{GW} sources \\
        $d_L$ & Luminosity distance of the \ac{GW} sources \\
        $m^z_1$ & Primary mass of the \ac{BBH} sources in the detector frame \\
        $Y_t$ & detector-frame primary mass posterior probability distribution samples of the $t$-th \ac{GW} event \\
        $\mathbf{Y}$ & Set of detector-frame primary mass posterior probability distribution samples of all \ac{GW} events \\
        $\Theta_i$ & Hyperparameters of the \ac{(H)DPGMM} \\
        $\mathbf{\Theta}$ & Set of hyperparameters of the \ac{(H)DPGMM} \\
        $p(m^z_1|\mathbf{\Theta})$ & Detector-frame observed primary mass population distribution reconstructed with the \ac{(H)DPGMM} \\
        $\Lambda$ & Source-frame population model \\
        $\Omega$ & Cosmological model \\
        $p(m_1|\Lambda)$ & Source-frame primary mass population distribution assuming the population model $\Lambda$ \\
        $p(m^z_1|\Lambda, \Omega)$ & Detector-frame primary mass population distribution assuming the population model $\Lambda$ and cosmological model $\Omega$ \\
        $\mathrm{det}$ & detectability of the \ac{GW} events \\
        \botrule
    \end{tabular}
    \label{tab:notation}
\end{table}

@@

In this example, instead of transforming the source-frame population model to the detector frame for each $H_0$ and performing a grid search, we define a function that calculates $d_\mathrm{JS}$ for a given set of the free parameters.
We then use the optimization algorithm mentioned in Sec.~\ref{sec:remapping} to find the optimized free parameters that minimize $d_\mathrm{JS}$.
The optimization method requires bounds for the free parameters, which we show in Table~\ref{tab:bounds}.
\begin{table}
    \caption{Free parameter bounds for the joint inference presented in Section~\ref{sec:inference_multi}.}
    \begin{tabularx}{\linewidth}{>{\centering\arraybackslash}X >{\centering\arraybackslash}X}
        \toprule
        Parameter & Bounds \\
        \midrule
        $H_0$ & $[1, 350]$ \\
        $\alpha$ & $[1.01, 15]$ \\
        $\mu$ & $[0.01, 70]$ \\
        $\sigma$ & $[0.01, 60]$ \\
        \botrule
    \end{tabularx}
    \label{tab:bounds}
\end{table}
The initial guesses for the optimization are chosen randomly from a uniform distribution within the bounds.
The result of the remapping is shown in Fig.~\ref{fig:simulation_result_multi}.
\begin{figure}
    \script{plot_simulation_result_multi.py}
    \includegraphics[width=\linewidth]{figures/simulation_result_multi.pdf}
    \caption{
        Optimized parameter distribution for the mock data inference.
        Simulated values are marked in red.
        The shaded regions represent the 39.35\% and 90\% credible intervals.
    }
    \label{fig:simulation_result_multi}
\end{figure}
The recovered distribution, whose uncertainty budget comes from the remapping of the individual-\ac{PE} uncertainty in the non-parametric reconstruction, is consistent with the simulated values.

@@

Next, we chose the \textsc{power law + peak} model as the source-frame population model $\Lambda$ and the $\Lambda$CDM model as the cosmological model $\Omega$.
For the redshift population model, instead of the uniform distribution over comoving volume and source-frame time, we used the \textsc{power law} redshift evolution model described in Eq.~8 of \cite{KAGRA:2021duu}.
Similar to Sec.~\ref{sec:inference_multi}, we fixed some parameters of the population model and the cosmological model.
For the cosmological model, we fixed the parameters to the results from \cite{Planck:2018vyg} except $H_0$.
For the population model, we fixed $\alpha$, $w$, $m_\mathrm{min}$, $m_\mathrm{max}$, and $\delta$ to the results from \cite{LIGOScientific:2021aug}.
Note that the fixed parameters are not necessarily the true values, but they can be considered as a reasonable guess for the parameters.
The fixed parameters of the population model are shown in Table~\ref{tab:fixed_parameters}.
\begin{table}
    \caption{Fixed parameters of the $\Lambda$CDM model (left) and the \textsc{power law + peak} model (right).}
    \begin{tabularx}{\linewidth}{>{\centering\arraybackslash}X >{\centering\arraybackslash}X | >{\centering\arraybackslash}X >{\centering\arraybackslash}X}
        \toprule
        Parameter & Value & Parameter & Value \\
        \midrule
         & & $w$ & $0.024$ \\
        $\Omega_m$ & $0.315$ & $\alpha$ & $4.2\ \mathrm{M}_\odot$ \\
        $\Omega_\Lambda$ & $0.685$ & $m_\mathrm{min}$ & $5\ \mathrm{M}_\odot$ \\
        $w$ & $-1$ & $m_\mathrm{max}$ & $109\ \mathrm{M}_\odot$ \\
         & & $\delta$ & $4.9\ \mathrm{M}_\odot$ \\
        \botrule
    \end{tabularx}
    \label{tab:fixed_parameters}
\end{table}

We then perform the remaining steps in the same way as Sec.~\ref{sec:inference_multi}.
We set the bounds for the power-law index of the redshift population model $\kappa$ to $[-100, 100]$, and the bounds for other free parameters are the same as in Table~\ref{tab:bounds}.
The result of the remapping is shown in Fig.~\ref{fig:real_result_multi}, and it is consistent with the posterior distribution reported in \cite{LIGOScientific:2021aug}.
\begin{figure}
    \script{plot_real_result_multi.py}
    \includegraphics[width=\linewidth]{figures/real_result_multi.pdf}
    \caption{
        Optimized parameter distribution for the real data inference.
        The shaded regions represent the 39.35\% and 90\% credible intervals.
    }
    \label{fig:real_result_multi}
\end{figure}
We find \variable{output/real_H0.txt}, (68\% credible interval around the median). Nonetheless, Figure~\ref{fig:real_result_multi} shows a peak in the marginalized posterior distribution of $H_0$ around $140\, \mathrm{km/s/Mpc}$, with a width of around $30-35\, \mathrm{km/s/Mpc}$ roughly consistent with the uncertainty reported in \cite{LIGOScientific:2021aug} ($H_0 = 50^{+37}_{-30} \, \mathrm{km/s/Mpc}$).

The reason for the peak differing from the one reported in \cite{LIGOScientific:2021aug}, which is around $50\, \mathrm{km/s/Mpc}$, can be attributed to two main factors.
First, the choice of the Jensen-Shannon distance as a measure of ``closeness'' between the non-parametric reconstruction and the predicted population is an arbitrary choice and, despite being a commonly used metric, does not stem from a specific derivation and/or assumption on the process that generated the observations.
As we stated in Section~\ref{sec:remapping}, the output of our method is not a posterior distribution in the strict sense due to this choice: the derivation of a well-grounded metric to connect non-parametric and parametric distributions will be discussed in a future paper.
Another possibility is the approximation made while including the selection effects: as discussed in \cite{Essick:2023upv}, this choice can lead, to some extent, to a bias in the reconstruction, and thus be the reason why the peak of the posterior distribution for $H_0$ is shifted.
Accounting for selection effects while reconstructing the observed distribution is in principle possible -- although not explored in the currently available literature -- and will be, as well, investigated in the future.
Despite these challenges, our method provides a foundation for exploring different models efficiently and can significantly aid in the analysis of large datasets expected in the near future.
